An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-08-22 EconGrowth2.Rnw
Added notes about non-convergence.
Insert TEST text
Testing
Intro and Refs
Added Introduction, incl Lit Review as well as References at end.

Big picture
Why do we do this? Want to link energy and economics on national level 
and in a modeled approach,informing energy policy during energy transitions.
Usually macro energy economic models rely on PF with energy included. 
Key papers using CES in energy economics incl high impact such as 
Stern, Gerlagh, Tol etc. 
Cite in introduction met energy economists using PF approaches, specially CES.
But, which form to choose and using which parameters? 
Literature is inconclusive, very little actual empirical testing of such functions. 
Mostly CES in CGE models used by Wbank and others. 
CD adviced to be used by CBO. 
How do we select a functional form and what are the parameters to be used? On what basis do we select?

@@

Thoughts on literature review and contextualisation of our work

Imagine ourselves walking into a conversation on the role of energy in the economy.

1. Listen to what others say

Identify key issues and important voices

\item{energy has become an important strategic issue for economic growth, shocks in 1970s, spikes in energy prices generated renewed interest (IEA, IMF), supported by analysis on more structural depletion and technology race over time (Hamilton), cite reviews on energy and economics (Stern,...), cite articles that show close correlations energy use and economic growth (Cleveland etc)}

\item{economic theoretical frameworks adjusted: energy theory of value; Energy fundamental in biophysical economics (Cleveland, Hall), energy as one cost in production set; energy plays a role in production (extended Solow-Swan or CES (who introduced CES first) framework)}

\item{how to account for energy on a macro economic modeling? taken up in economic growth theory by (1) including energy in PF (stern, froling, tahvonen, Gerlagh, Tol, Kemfert, FEEM crowd, Cooper, Nel) or (2) new thermodynamically rooted Function (kümmel, lindenberger, ayres, warr). Mention energy in endogenous growth theory models (smulders, Mulder, vd Bergh) in SD modelling}

\item{countries are going through different stages in their development, and the role of energy in such phases is now started to be explored (frohling, tahvonen, dalgaard)}

\item{some have started to do work on empirical/statistical testing of PF with energy (Shen, walley, Serrenho,) and some have started doing comparative analysis of PF CD vs CES (miller)}

2. Find our own voice

What is our own argument?

We argue that macroeconomic modelling is a reality (being used for policy analysis), 
but needs to be used in a very careful way. 
Throughout decades, there are enough signals that 
macroeconomic growth theory has serious limitations. 
Not revisiting but refer to reviews (mishra, others).

We argue that the basis on which choice version of energy-augmented macroeconomic model 
and parameters are made, needs more detailed work, 
specifically detailed statistical testing and/or data analysis, including
the quantification of energy ($q$, $x$, or $u$).

We also argue that different types of countries may need different PF parameters 
to be used in modelling approaches, 
leave possibility open for different explanations of growth 
in different countries and at different stages of their development.

We present a comparative analysis among countries and of commonly used approaches in PF, 
focussing on the insights one may drawn on a detailed comparative analysis, 
rather then run with one selected PF.

What is general tone of conversation? Controversial, established? Cross-disciplinary?
We provide insights from cross-disciplinary work among statistics, engineering,
and environmental economics. This is a new approach to issues that have been 
discussed in recent years in the literature.

We are taking issue with untested (statistically) applications of functional forms 
of energy-augmented PF across different countries, with different energy quantifications. 
Also taking issue with idea that importance of energy in growth 
also means it is necessary to include it as a factor in PF (unnecessary vs. unimportant). 
Opens up a argument for the need of richer than just aggregate modelling approaches.

Who makes arguments we want to amplify?

Amplify voices who claim energy is important in understanding growth in economies (Stern, Tol, Gerlagh)

Amplify those who call for caution in applying aggregate PF by just adding energy to it. (who?)

Amplify those who call for testing (statistically) the benefit of energy-augmentation. (shen/whalley)

Amplify comparative analysis of role of energy between countries indifferent stages of development. (who?)

What established sources we build on? Question? Contradict? In our writing.

Build on articles using R to test econ growth PF, 
question what is reported when doing these tests (shen/whalley and others who have used micEconCES). 
Critique literature that calls for uncritical energy-augmentation of PF, 
without testing whether that make sense at all.

Does analysis differ from others in the conversation?

Yes, differs as follows:
\begin{itemize}
\item{if one do these test, report on all $\delta$, $\sigma$, $\gamma$, and $\lambda$ 
values as well as sse. 
Good fits not evidence of best values, must report uncertainties as well.}
\item{What motivates nesting of CES? Not well behaved $\rho$ results, driving $\sigma$s.
Different type of CES with parsimonious factors, sigmas can be generated but not that easy (idea Randy).
If you run all 3 nesting options, you can identify elasticities of all three factors in relation to each 
other. Furthermore, you can determine which two factors are driving overall behavior in the aggregate function.}
\item{necessary of different factors in PF to explain growth for countries in diff. development status}
\end{itemize}

@@

@@

>>>>

Introduction

[1. Listen to what others say
Identify key issues and important voices]

The available of adequate energy at affordable costs has become an important strategic issue in recent years.  Despite a slowdown in economic growth worldwide, longer-term trends in energy use remain strongly upwards, as the dynamics of energy markets are increasingly determined by emerging markets, notably China, India and the Middle East (IEA, 2012).  A large body of literature points towards the fact that economic growth and energy consumption are closely correlated and for a diverse set of countries such as the United States (Stern, 2000; Cleveland et al 1984), India (Paul & Bhattacharya 2004), China (Wang et al 2011), South Africa (Menyah & Wolde-Rufael, 2010; Odhiambo, 2009) and sixteen Asian countries (Lee & Chang, 2008) to name a few, although the causal relationship between energy consumption and economic growth is still controversial (Chen et al 2012; Ozturk, 2009).  Ozturk (2009:347) in a survey of the literature on the topic, concluded that there is “… no consensus neither on the existence nor on the direction of causality between these variables in the literature”, but do point out that “…causality is running from electricity consumption to economic growth in the most of these studies [surveyed]”.

Those that have accepted that energy use is closely correlated to economic performance in one form or another, have expressed further concerns on the negative effects of rising energy prices and energy depletion effects on current and future economic growth.  Benes et al (2012), incorporating both views on geological physical constraints and technological solutions to oil production, suggest a near doubling of the real price of over the next decade.  Kumhof & Muir (2012) point out that for larger oil supply shocks, assuming production functions where oil plays a prominent role as well as reduced possibilities for substitution away from oil, significant impacts on GDP growth could occur.  Nel and Van Zyl (2010) argue for the “…vital importance of energy security…” on the basis of a model based on an explicit energy-based formulation of economic growth.  Hamilton (2012:30) argues that “[c]oping with a final peak in world oil production…have much in common with previous historical episodes that resulted from temporary geopolitical conflict, being associated with significant declines in employment and output. If the future decades look like the last 5 years, we are in for a rough time.”

The current trends in energy production and consumption and the possible dire impacts on the world economy as argued by above-mentioned scholars have further raised the question whether and how to including appropriate energy indicators in economic growth modeling.  Stern (2011) provided a useful overview on how energy is dealt with in mainstream economic, resource economic and ecological economic models of growth.  Stern points out that energy is usually not included as a factor of production in mainstream economic growth models, that energy is seen as the central driver of growth in ecological economic (incl. biophysical economic) models, and that energy is included in resource economic models, but remain isolated within this sub-discipline (2011:26).  To break through this theoretical impasse, Stern proposed a synthesized CES production function where the constraints or abundance of energy determines whether to include energy explicitly as a factor of production or not.  On the basis of this model, Stern (2011:45) points out that energy availability plays “…a key role in enabling growth” even given the evidence of decreasing energy used per unit of output.  He concludes that, based on theory and time-series results the inclusion of energy in the production function approach to economic growth is warranted, and that further clarification is needed on the role of energy in economic growth.  The inclusion of energy as a factor of production in CES and Cobb-Douglas production functions, is generally accepted by most resource and energy economists and economic growth modelers (see Smulders & De Nooij 2003; Van der Zwaan et al 2002; Kemfert, 1998).  Researchers interested in explaining long-term economic growth and possible constraints on growth have started to include energy as an explicit production factor (Fröling, 2011; Tahvonen & Salo, 2001).  One alternative approach, attempting to be “…more consistent with physical laws…” (Ayres, 2001), and classified by Stern as an ecological economic model of growth is the LINEX production function approach, combining capital (non ICT and ICT), labour and useful work, but with the difference that technological change only affects the amount of useful work done per joule of exergy (Ayres & Warr 2005).

The unsatisfactorily situation is that the choice of a production function and the parameters used to populate these functions in economic growth modeling will affect the results in a significant way.  This leaves far too much scope to misuse model specifications and the selective choice of parameters to support certain theories rather then empirically test these theories.  For example, Lecca et al (2010) pointed out that the modellers’ choice on the structure of a production function reflects different views on the elasticity between different inputs and “…the point at which energy enters the [nested CES] production function has a significant impact on CGE [computable general equilibrium] model results.” [brackets added].  Duffy and Papageorgiou (2000) rejected the Cobb-Douglas specification altogether finding that capital and labour are more substitutable in richer countries and less substitutable in poorer countries then what a Cobb-Douglas specification would imply.  Miller (2008), however, points to the “…reasonably accurate long-term economic forecasts”, achieved with the use of Cobb-Douglas production functions in economic models.  Shen and Whalley (2013) is one recent example where empirical analysis on the estimated substitution elasticities is actually done before choosing any particular nested CES production function structure and as a result propose a different structure to the Chinese economy to the one usually used in macro-economic growth modeling.  Both the choice of a production function structure and the value of the parameters chosen drive model results and empirical and statistical testing of such approaches appear more and more prudent.

[2. Find our own voice
What is our own argument?]

We agree with those who have a serious concern with the reliability of the results from economic growth models which are often used to inform policy choices (Lecca et al 2010) and would like to add that the inclusion of energy would only further complicate issues.  We do except, however, that macroeconomic policymaking is needed and are increasingly informed by integrated energy-economic models, in a world where energy constraints are possible (Beckman et al 2011; Benes et al, 2010; Nel and van Zyl, 2010).  We support the call from Shen and Walley (2013) that the basis on which a modelling choice is made on the structure of the production function and the parameters used to populate this function, needs empirical testing, and we would like to add; including for the quantification of energy in various measures ($q$, $x$, or $u$).  Building on the argument of Stern (2011) that energy may or may not be required in production functions depending on its abundance or scarcity, we will further argue in this paper that different types of countries may need different factors of production to explain their respective growth paths.  This further leaves the possibility open for different explanations of growth in different countries and at different stages of their development, a central point by contributions made in long-wave economic growth modeling (Fröling, 2011; Jones, 2001; Galor and Weil, 2000).  To gain insights from different countries across different stages of development, we present a comparative analysis among the World Bank’s main categories of countries, namely Developed Countries (DCs), Newly Industrialized Countries (NICs), OPEC countries, and Least Developed Countries (LDCs).

Approach
Build on articles using R to test econ growth PF, 
question what is reported when doing these tests (shen/whalley and others who have used micEconCES).

Discussion and Insights

Does analysis differ from others in the conversation?
Yes, differs as follows:
\begin{itemize}
\item{if one do these test, report on all $\delta$, $\sigma$, $\gamma$, and $\lambda$ 
values as well as sse. 
Good fits not evidence of best values, must report uncertainties as well.}
\item{What motivates nesting of CES? Not well behaved $\rho$ results, driving $\sigma$s.
Different type of CES with parsimonious factors, sigmas can be generated but not that easy (idea Randy).
If you run all 3 nesting options, you can identify elasticities of all three factors in relation to each 
other. Furthermore, you can determine which two factors are driving overall behavior in the aggregate function.}
\item{necessary of different factors in PF to explain growth for countries in diff. development status}
\end{itemize}

Big picture
[insert test test]

Why do we do this? Want to link energy and economics on national level 
and in a modeled approach,informing energy policy during energy transitions.
Usually macro energy economic models rely on PF with energy included. 
Key papers using CES in energy economics incl high impact such as 
Stern, Gerlagh, Tol etc. 
Cite in introduction met energy economists using PF approaches, specially CES.
But, which form to choose and using which parameters? 
Literature is inconclusive, very little actual empirical testing of such functions. 
Mostly CES in CGE models used by Wbank and others. 
CD adviced to be used by CBO. 
How do we select a functional form and what are the parameters to be used? On what basis do we select?

@@

@@

**** Do we want to report on the fraction of non-convergence
reported? Maybe we should write a function that 
searches all the countries to find which 
country reports least convergence for a given model. ****

@@

[manually added]
References

Benes, J., Chauvet, M., Kamenik, O., Kumhof, M., Laxton, D., Mursula, S., & Selody, J. (2012). The Future of Oil: Geology versus Technology. IMF Working Papers, I12/109, 1–33.

Chen, P-Y., Chen, S-T., Chen, C-C. 2012. Energy consumption and economic growth – New evidence from meta analysis. Energy Policy, 44: 245-255. http://dx.doi.org/10.1016/j.enpol.2012.01.048

Cleveland, C., Costanza, R., Hall, C., & Kaufmann, R. (1984). Energy and the US economy: a biophysical perspective. INTERNATIONAL LIBRARY OF CRITICAL WRITINGS IN ECONOMICS, 75, 295–302.

Duffy, J., & Papageorgiou, C. (2000). A Cross-Country Empirical Investigation of the Aggregate Production Function Specification. Journal of Economic Growth, 5(1), 87–120. doi:10.1023/A:1009830421147

Fröling, M. (2011). Energy use, population and growth, 1800–1970. Journal of Population Economics, 24(3), 1133–1163.

Galor, O., & Weil, D. N. (2000). Population,Technology, and Growth: From Malthusian Stagnation to the Demographic Transition and Beyond. American Economic Review, 90(4), 806–828.

Hamilton, J. (2012). Oil Prices, exhaustible resources, and economic growth.  Handbook of Energy and Climate Change

International Energy Agency (IEA), 2012. World Energy Outlook 2012. IEA: Paris, France.

Jones, C. (2001). Was an industrial revolution inevitable? Economic growth over the very long run. Advances in Macroeconomics, 1(2), 1–43.

Kemfert, C. (1998). Estimated substitution elasticities of a nested CES production function approach for Germany. Energy Economics, 20(3), 249–264.

Kumhof, M., & Muir, D. (2012). Oil and the World Economy: Some Possible Futures. IMF Working Papers, 12/256(October, 1), 1–31.

Lee, C-C. & Chang, C-P. Energy consumption and economic growth in Asian economies: A more comprehensive analysis using panel data. Resource and Energy Economics, 30(1): 50-65.

Lecca, P., Swales, K., & Turner, K. (2011). An investigation of issues relating to where energy should enter the production function. Economic Modelling, 28(6), 2832–2841.
